A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
More work on the methods section

As is the case with cross-sectional data, it is possible to test the residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. A sp

>>>>

As is the case with cross-sectional data, it is possible to test the residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. Spatial analytical approaches were used to understand contagion-difussion processes in the case of infectious disease in general [e.g., @Cliff1998detecting] and the 2003-2004 SARS outbreak in particular [e.g., @Meng2005understanding; @Cao2010spatio]. While we are mindful of the same caveat that the novel SARS-CoV2 may not follow the patterns of previous diseases, there is still evidence from the United States that COVID-19 displays spatial patterns that are consistent with a diffusion process [@Desjardins2020rapid]. For this reason, the spatial lag model is appropriate to model incidence of COVID-19 geographically, since it accounts for potential spatial patterns that result from a process of contagion, as explained next.

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
More work on the methods section

The first reported case of COVID-19 in Spain was on January 31th, when a
German tourist in the Canary Islands tested positive for the virus.
However, it was still a few weeks before the first domestic case was
reported, on February 27th in Sevilla province (Andalusia). In a short
period of time, after this relatively slow start, the outbreak flared.
By March 11th the World Health Organization (WHO) declared COVID-19
officially a pandemic. This declaration marked a turning point for the
public in Spain too. As of March 13th, the number of cases of COVID-19
reported in Spain was 4,473, with a majority of cases (1,990)
concentrated in Madrid: these numbers were at the time the worst
outbreak in Europe, after Italy. In response to the situation, on March
13th the Spanish National Government declared a state of emergency. As
part of the state of emergency restrictions to most activities were
imposed, with the exception of essential services (e.g.~food, health)
and some economic subsectors of industry and construction. A few days
later, on March 17th, Spain closed its lands borders to allow entry only
to returnee nationals and permanent residents. The lockdown was further
hardened between March 30th and April 12th (including the Easter weekend
of April 10th-12th) and during this period only essential activities
were allowed. During this period, there was a dramatic reduction in
overall mobility, both within provinces
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}
as between
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

>>>>

The first reported case of COVID-19 in Spain was on January 31th, when a
German tourist in the Canary Islands tested positive for the virus.
However, it was still a few weeks before the first domestic case was
reported, on February 27th in Sevilla province (Andalusia). In a short
period of time, after this relatively slow start, the outbreak flared.
By March 11th the World Health Organization (WHO) declared COVID-19
officially a pandemic. This declaration marked a turning point for the
public in Spain too. As of March 13th, the number of cases of COVID-19
reported in Spain was 4,473, with a majority of cases (1,990)
concentrated in Madrid: these numbers were at the time the worst
outbreak in Europe, after Italy. In response to the situation, on March
13th the Spanish National Government declared a state of emergency, to
go into effect on Saturday March 14th. As part of the state of emergency
restrictions to most activities were imposed, with the exception of
essential services (e.g.~food, health) and some economic subsectors of
industry and construction. A few days later, on March 17th, Spain closed
its lands borders to allow entry only to returnee nationals and
permanent residents. The lockdown was further hardened between March
30th and April 12th (including the Easter weekend of April 10th-12th)
and during this period only essential activities were allowed. During
this period, there was a dramatic reduction in overall mobility, both
within provinces as between
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.
